Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-07-29 Manuscript/MPESubmission/Revision/response.tex

\item{\textcolor{blue}{\textbf{line 173 to 176:} need some justification for why the substitution rate with alpha shape of 0.5  will avoid homoplasy. Yang 1996 is cited but clarify the justification of what is 'too much' homoplasy. }}

We changed the sentence to:
"In practice, such a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal (Hassanin et al., 1998; Estoup et al., 2002).%\citep{Hassanin1998611,EstoupHomoplasy}.
Also, we chose this low value to be consistent with our protocol for simulating morphological characters (see below)." (line @@@).

\item{\textcolor{blue}{\textbf{Line 176 to 180:} could be streamlined. It is not clear what is meant by 'no special assumption about how the characters evolved', what constitutes a special assumption? Could potentially be reduced to something like 'this model and these parameter settings strike a balance between realism for empirical datasets (citations) and parameter richness with more complex models (e.g., GTR, multiple partitions with independent models), making them more suitable for our computational limitations'. Just for example.}}

We modified the sentence as follows:
"We selected this model and parameters settings to be a compromise between realism from empirical datasets (e.g. Kelly et al., 2014) and models with few parameters (cf. Springer et. al., 2012) to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years)."

@@

We rephrased the section underlined by the reviewer as follows:

"We used low evolutionary rate parameters to be consistent with the molecular rates parameters and to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal (Wright and Hillis, 2014). Note that Wright and Hillis (2014) have shown that low morphological rates (< 0.5) increases variance in topological error but we discarded simulations with such topological error by selecting only matrices with a “fair” phylogenetic signal (see Estimating phylogenies section below; Zander, 2004)."

\item{\textcolor{blue}{\textbf{line 201:} change 'get' for 'obtain' or something less colloquial.}}

We modified this whole section following reviewer 2 comment 2.

@@

\item{\textcolor{blue}{\textbf{line 231-235:} streamline, e.g., 'The minimal dataset included in this study was 5\% of the morphological matrix for any taxon'.}}

We modified the sentence as follows: "To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon." (line @@@).

\item{\textcolor{blue}{\textbf{line 237-246:} I understand the rational presented by the authors for comparing missing data trees to the best tree inferred from max. likelihood/Bayesian techniques rather than compare missing data trees to the true tree used to simulate the data. I still feel it would be an important contribution to this paper, however, to compare the missing data trees to the true tree to evaluate the effects of missing data on the tree inference. Also it may lend insights to whether the parameter settings used in the simulations and in the analyses were adequate for generating datasets and inferring the true tree, or something close to it. }}

We do agree with the point raised with both reviewers on comparing the true tree to the best/missing-data trees.
As reviewer 1 states: "[this analysis can give] insights to whether the parameter settings used in the simulations and in the analyses were adequate for generating datasets and inferring the true tree, or something close to it.".
We've added the results of these comparisons to the supplementary materials (Appendix C - figures C5 and C6).
However, even though we agree with the comments of both reviewer that knowing the difference between the true and best/missing-data trees is important, we do not think that it "adds an additional layer of complexity that can't be accounted for via simulation" (reviewer 2, comment 1).
In this simulations, we had to make a choice between (1) simulating larger matrices but with less chains (i.e. less variance) or (2) smaller matrices with more variance.
We deliberately made the second choice because our main question was "what is the effect of our parameters on recovering the topology".
In fact, regarding our problematic, we maintain that comparing the best tree to the missing-data trees is the best approach to answer this specific question for two reasons:
\begin{enumerate}
\item{The true tree is always unknown:} as we state in the manuscript: "In real world situations, the ``true'' tree is never available" (line 242 in the first manuscript and @@@ in the revision).
\item{The true tree is just the starting point of the simulations:} in fact, we use the true tree only as a starting point for our simulations. Once generated, the true tree is fed to a cascade of algorithms and softwares (i.e. seqGen\{phyclust\}, rTraitDisc\{ape\}, RAxML, MrBayes) described in section 2.1 and 2.3 in order to get the best tree and the missing-data trees. Comparing the latests to the true tree will answer to the problematic "what is the effect of our missing-data parameters and the cascade of algorithms and softwares on recovering the true topology?".
\end{enumerate}
Therefore we added the second justification to the main text (see below this response) and added these analysis (true vs. best/missing-data trees) to the supplementary for potential readers that might be interested by this specific problematic.

@@

\item{\textcolor{blue}{\textbf{Appendix D:} I find the online code repository thorough, but the repository is loaded with many files, especially for graphing. If it is possible to condense these files it may be easier for a reader to duplicate the study.}}

%Working on it.

\end{enumerate}

\section{Reviewer 2:}
Note that we split the second comment from reviewer 2 in three different comments (comments 2,3 and 4 below) to improve the readability of our response to the reviewers.

### 2015-07-29 Manuscript/MPESubmission/Revision/OnlineAppendices-LaTeXSuppFiles/AppendixB.tex
True trees analysis

>>>>

\subsection{Differences between the ``true'' and the ``best'' trees.}
We compared every ``true'' trees (i.e. the trees \textbf{used to create} the ``complete'' matrices) to the ``best'' trees (i.e. the trees \textbf{inferred from} the ``complete'' matrices) in figure ~\ref{Fig_Supp_True_Best}.
Note that the difference between the ``true'' and the ``best'' trees represents the effect of the parameters choice and the algorithms used to create the ``complete'' matrix as well the as the capacity of RAxML and MrBayes to infer phylogenies from this particular matrices (i.e. small sized and generated using specific algorithms).
This does not affect, however, the results of our analysis since we deliberately compared the the ``missing-data'' trees to the ``best'' tree rather than to the ``true'' tree.

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{SupplementaryFigures/True_vs_Best_trees.pdf}
    \caption{Pairwise comparisons between the 50 ``true'' trees and the 50 ``best'' trees from the Maximum Likelihood and Bayesian inference methods. The horizontal blue and red lines represents respectively the 95\% and the 50\% confidence intervals.} 
\label{Fig_Supp_True_Best} 
\end{figure}

### 2015-07-29 Manuscript/MPESubmission/Revision/OnlineAppendices-LaTeXSuppFiles/AppendixC.tex
Updated figure 4 (with Tr below 0)
True trees analysis

>>>>

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{SupplementaryFigures/Singles_True.pdf}
    \caption{Effect of increasing missing data on recovering the ``true'' tree topology (the tree used for starting our simulations) for the Maximum Likelihood trees (black) and Bayesian consensus trees (grey). The x axis shows the percentage of missing data from 0\% (white) to 75\% (black) for the two parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and number of characters from 100 to 25 for the parameter $N_{C}$ (lower line). Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds metric (upper row) and Normalised Triplets metric (lower row). The graph shows the modal value (points), and the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.} 
\label{Fig_Supp_Singles_true} 
\end{figure}

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{SupplementaryFigures/Treesets_True.pdf}
    \caption{Effect of increasing missing data on topological recovering the ``true'' tree topology (the tree used for starting our simulations) for the Maximum Likelihood Bootstrap trees (black) and Bayesian posterior tree distribution (grey). The x axis shows the percentage of missing data from 0\% (white) to 75\% (black) for the two parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and number of characters from 100 to 25 for the parameter $N_{C}$ (lower line). Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds metric (upper row) and Normalised Triplets metric (lower row). The graph shows the modal value (points), and the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.} 
\label{Fig_Supp_Treesets_true} 
\end{figure}

### 2015-07-29 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
True trees analysis

% TG: this part is entirely re-written.
To explore the effect of missing data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
We removed morphological following three main data incompleteness parameters:
\begin{enumerate}
\item{The proportion of missing living taxa ($M_{L}$).}
This first missing-data parameter corresponds to the proportion of living taxa with no morphological data.
It represents the number of living taxa that are present in the matrix but have only molecular data available.
This reflects the fact that, because of the increasing facility of collecting molecular data, morphological data for living species is rarely collected \citep{GuillermeCooperMissing}. % TG: see what I did here? ;)
Therefore, many living species will have only molecular data available.
In practice, we removed all the morphological data from randomly chosen living with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.
\item{The proportion of missing data in the fossil record ($M_{F}$).}
This missing data parameter represents the completeness of the fossil record.
In fact, due to preservation biases, missing data for fossil taxa are common \citep{sansomfossilization2013}. 
In practice, we randomly removed a proportion of data among the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
\item{The number of morphological characters for both living and fossil taxa ($N_{C}$).}
This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix.
Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large. % TG: this is word for word what the reviewer wrote. I didn't like doing that first but after seeing the Ineterface Focus review and how they copy pasted the discussion suggestion, I think it's really nice for the reviewer to see a bit of it's work directly in the paper (at least it was for me!).
Therefore, this parameter correspond to the overall number of characters available for both living and fossil taxa.
In practice, we randomly removed whole characters from the morphological matrix shrinking it to: 100, 90, 75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).
\end{enumerate}

% TG: this part is nearly exactly the same as before.
In practice, each parameter represents a different way of removing data from the morphological part of the matrix: $M_L$ removes entire rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $N_C$ removes columns across both living and fossil taxa's data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.

% TG: and this one is slightly modified only.
We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data''matrices.
Note that one of these combinations ($M_{L}$=0\%; $M_{F}$=0\% and $N_{C}$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).

>>>>

% TG: this part is entirely re-written.
\hl{To explore the effect of missing data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
We removed morphological following three main data incompleteness parameters:}
\begin{enumerate}
\item{\hl{The proportion of missing living taxa ($M_{L}$).}}
\hl{This first missing-data parameter corresponds to the proportion of living taxa with no morphological data.
It represents the number of living taxa that are present in the matrix but have only molecular data available.
This reflects the fact that, because of the increasing facility of collecting molecular data, morphological data for living species is rarely collected} \citep{GuillermeCooperMissing}. % TG: see what I did here? ;)
\hl{Therefore, many living species will have only molecular data available.
In practice, we removed all the morphological data from randomly chosen living with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.}
\item{\hl{The proportion of missing data in the fossil record ($M_{F}$).}}
\hl{This missing data parameter represents the completeness of the fossil record.
In fact, due to preservation biases, missing data for fossil taxa are common }\citep{sansomfossilization2013}. 
\hl{In practice, we randomly removed a proportion of data among the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.}
\item{\hl{The number of morphological characters for both living and fossil taxa ($N_{C}$).}}
\hl{This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix.
Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large. % TG: this is word for word what the reviewer wrote. I didn't like doing that first but after seeing the Ineterface Focus review and how they copy pasted the discussion suggestion, I think it's really nice for the reviewer to see a bit of it's work directly in the paper (at least it was for me!).
Therefore, this parameter correspond to the overall number of characters available for both living and fossil taxa.
In practice, we randomly removed whole characters from the morphological matrix shrinking it to: 100, 90, 75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).}
\end{enumerate}

% TG: this part is nearly exactly the same as before.
In practice, each parameter represents a different way of removing data from the \hl{morphological part of} the matrix: $M_L$ removes \hl{entire} rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $N_C$ removes columns across both living and fossil taxa's data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.

% TG: and this one is slightly modified only.
\hl{We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data''matrices.
Note that one of these combinations ($M_{L}$=0\%; $M_{F}$=0\% and $N_{C}$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).}
